---
title: "R_functional_script"
author: "Sam Westreich, stwestreich@ucdavis.edu, github.com/transcript"
date: "June 24, 2015"
output: html_document
---

This is a script for processing output files from MG-RAST, converting functional output data into a graph.  This program creates a barplot of activity within a sample grouped by protein function, both by relative and by absolute activity.

Necessary packages to include:

```{r packages, results='hide'}
library(DESeq)
library(DESeq2)
library(ggplot2)
library(gridExtra)
library(scales)
library(reshape)
library(knitr)
library(plyr)
```

Before starting this script, please adjust the following parameters.  The working directory should contain both the experimental sample files (designated by "exp_<filename>") and the control sample files (designated by "control_<filename>").  

```{r working_directory}
setwd("~/Desktop/Projects/Lab Stuff/SAMSA_pipeline_v1/public_output_files/public_files_ready_for_R/function/")
```

Now, we read in the files from the working directory as specified above.  The file should be of a format such as:
"control_RefSeq_function_identifier-name.tab.output".
Note that if the identifier name or ID of the file is not specified as such, the script will not select the correct section of the file name to use as an ID.

```{r file_listing}
control_files <- list.files(
    pattern = "control_*", full.names = T, recursive = FALSE)
control_names = ""
for (name in control_files) {
  control_names <- c(control_names, unlist(strsplit(name, split='_', fixed=TRUE))[2])} #change the "2" to a different part of the name if not following standard naming conventions as above
control_names <- control_names[-1]
control_names_trimmed = ""
for (name in control_names) {
  control_names_trimmed <- c(control_names_trimmed, unlist(strsplit(name, split='.', fixed=TRUE))[1])}
control_names_trimmed <- control_names_trimmed[-1]
rm (control_names)

exp_files <- list.files(
    pattern = "experiment_*", full.names = T, recursive = FALSE)
exp_names = ""
for (name in exp_files) {
  exp_names <- c(exp_names, unlist(strsplit(name, split='_', fixed=TRUE))[2])} #again, change 2 if non-standard file names are used
exp_names <- exp_names[-1]
exp_names_trimmed = ""
for (name in exp_names) {
  exp_names_trimmed <- c(exp_names_trimmed, unlist(strsplit(name, split='.', fixed=TRUE))[1])}
exp_names_trimmed <- exp_names_trimmed[-1]
rm (exp_names, name)
```

Next, we will load these files in as two tables, one for the experimental data and one for the control data.

```{r open_tables}
# loading the control table
y <- 0
for (x in control_files) {
  y <- y + 1
  if (y == 1) {
    control_table <- read.table(file = x, header=F, quote = "", sep = "\t", fill = TRUE)
    colnames(control_table) = c(x, "V3", "md5")
    control_table <- control_table[,c(3,1,2)] 
    control_table <- control_table[c(1:1000),] }
  if (y > 1) {
    temp_table <- read.table(file = x, header = F, quote = "", sep = "\t", fill = TRUE)
    colnames(temp_table) = c(x, "V3", "md5")  
    temp_table <- temp_table[c(1:1000),]  }
  if (y > 1) {
    print (x)
    control_table <- merge(control_table, temp_table[,c(2,3)], by = "V3", all.x = T)  }
}
control_table[is.na(control_table)] <- 0
rownames(control_table) = control_table$V3
control_table_trimmed <- control_table[,-3, drop = FALSE]

# loading the exp table
y <- 0
for (x in exp_files) {
  y <- y + 1
  if (y == 1) {
    exp_table <- read.table(file = x, header=F, quote = "", sep = "\t", fill = TRUE)
    colnames(exp_table) = c("DELETE", x, "V3", "md5")
    exp_table <- exp_table[,c(4,2,3)]  }
  if (y > 1) {
    temp_table <- read.table(file = x, header = F, quote = "", sep = "\t", fill = TRUE)
    colnames(temp_table) = c("DELETE", x, "V3", "md5")  }
  if (y > 1) {
    print (x)
    exp_table <- merge(exp_table, temp_table[,c(2,3)], by = "V3", all.x = T)  }
}
exp_table[is.na(exp_table)] <- 0
rownames(exp_table) = exp_table$V3
exp_table_trimmed <- exp_table[,-3, drop = FALSE]
rm(control_table, exp_table, x, y)
```

Now, we simplify the names of the columns of these two tables, using the ID names scrubbed from the filenames above in chunk 3, file_listing.  If this step is failing, please note the proper naming of the files as mentioned above chunk 3.

```{r column_names}
colnames(control_table_trimmed) = c("md5", control_names_trimmed)
colnames(exp_table_trimmed) = c("md5", exp_names_trimmed)
```

******** GRAPH FORMATION ******** 
NOTE: If skipping directly to DESeq analysis of the results, proceed instead to line 19X.

The next step is to remove the hypothetical or predicted proteins.  Note that this significantly reduces the total amount of data, but provides much more clarity when examining end results.

```{r hypothetical_removal}
control_to_be_removed <- control_table_trimmed[grep("hypothetical", rownames(control_table_trimmed)),]
control_table_filtered <- control_table_trimmed[ !(rownames(control_table_trimmed) %in% rownames(control_to_be_removed)),]
control_to_be_removed <- control_table_trimmed[grep("predicted", rownames(control_table_trimmed)),]
control_table_filtered <- control_table_filtered[ !(rownames(control_table_filtered) %in% rownames(control_to_be_removed)),]

exp_to_be_removed <- exp_table_trimmed[grep("hypothetical", rownames(exp_table_trimmed)),]
exp_table_filtered <- exp_table_trimmed[ !(rownames(exp_table_trimmed) %in% rownames(exp_to_be_removed)),]
exp_to_be_removed <- exp_table_trimmed[grep("predicted", rownames(exp_table_trimmed)),]
exp_table_filtered <- exp_table_filtered[ !(rownames(exp_table_filtered) %in% rownames(exp_to_be_removed)),]
rm(control_to_be_removed, exp_to_be_removed)
```

We now want to combine the two tables (experimental and control) into a single table, designated as all_table.

```{r all_table_creation}
all_table <- control_table_filtered
all_table[, " "] <- 0
all_table <- merge(all_table, exp_table_filtered, by = c(0,1), all = TRUE) # this step may take a while, depending on total number of annotations
all_table[is.na(all_table)] <- 0
rownames(all_table) <- all_table[,2]
all_table <- all_table[,-2]
all_table[, "Total"] <- rowSums(all_table[,c(2:length(all_table))])
all_table <- all_table[ with(all_table, order(-Total)), ]
names(all_table)[names(all_table) == 'Row.names'] <- 'Function'

# let's save everything up to this point:
all_table_all <- all_table

# continue here to include only the top 30 most abundant functions.
all_table <- rbind(all_table[1:29,], all_table[nrow(all_table),])
all_table <- all_table[, -ncol(all_table)]
```

We now alter the format of the all_table using melt:

```{r merge_tables}
all_table_m <- melt(cbind(all_table, Function = rownames(all_table)), id.vars = c('Function'))
```

One difficulty with function exports from RefSeq is that there are often several different vernacular forms of each function, split by semicolon.  To simplify remaining functions, we here split by semicolon and merge functions with identical vernacular.

```{r function_simplifying}
all_table_m$Simplified_Function <- sapply(strsplit(as.character(all_table_m$Function), ";"), "[", 1)
all_table_m$Simplified_Function <- as.factor(all_table_m$Simplified_Function)

all_table_m2 <- all_table_m[c(4,2,3)]
all_table_m2 <- ddply(all_table_m2, c("Simplified_Function", "variable"), numcolwise(sum))
names(all_table_m2)[1] <- "Function"
all_table_m2$Function <- as.factor(all_table_m2$Function)
all_table_m3 <- all_table_m2[order(all_table_m2$variable, all_table_m2$Function),]
rm(all_table_m2, all_table_m)
```

Because normal ggplot colors tend to blend together too much when required to select 30 colors, a custom palette is specified here:

```{r palette}
CbPalette <- c("#a6cee3", "#1f78b4", "#b2df8a", "#33a02c",  "#fb9a99", "#e31a1c", "#fdbf6f", "#ff7f00", "#cab2d6", "#6a3d9a", "#ffff99", "#b15928", "#8dd3c7",  "#ffffb3",  "#bebada",  "#fb8072",  "#80b1d3",  "#fdb462",  "#b3de69", "#fccde5",  "#d9d9d9",  "#bc80bd",  "#ccebc5",  "#ffed6f", "#e41a1c",  "#377eb8", "#4daf4a",  "#984ea3",  "#ff7f00", "#ffff33",  "#a65628",  "#f781bf",  "#999999", "#000000")
```

Now, we create the graphs!  First, we create a stacked bar by percentage, allowing us to see relative activity composition by sample. 

```{r percentage_plot}
relative_ggplot <- ggplot(all_table_m3, aes(x = variable, y = value, fill = Function)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_fill_manual(values = CbPalette) +
  scale_y_continuous(labels = percent_format()) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(ncol=4)) +
  ggtitle("Top 30 protein functions by relative abundance") +
  xlab("Sample ID") + ylab("Relative activity of total sample")

relative_ggplot
```

Next, we create a stacked bar graph by raw counts, showing how total sample counts compare to each other.

```{r absolute_plot}
absolute_ggplot <- ggplot(all_table_m3, aes(x = variable, y = value, fill = Function)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = CbPalette) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(ncol=5)) +
  ggtitle("Top 30 protein functions by total abundance") +
  xlab("Sample ID") + ylab("Activity reads per sample")

absolute_ggplot
```

Finally, we display both of these graphs on a single display!

```{r graph_output}
grid.arrange(relative_ggplot, absolute_ggplot, ncol = 1)
```

******** DESEQ ANALYSIS ******** 

Now, let's run some DESeq analysis on the results:

```{r DESeq_analysis}
library(DESeq2)
complete_table <- merge(control_table_trimmed, exp_table_trimmed, by=c(0,1), all = TRUE)
complete_table[is.na(complete_table)] <- 1
rownames(complete_table) <- complete_table$md5
complete_table <- complete_table[,-2]

complete_table$Simplified_Function <- sapply(strsplit(as.character(complete_table$Row.names), ";"), "[", 1)
complete_table$Simplified_Function <- as.factor(complete_table$Simplified_Function)
complete_table <- complete_table[,c(ncol(complete_table), 1:ncol(complete_table))]
complete_table <- complete_table[,c(-2, -ncol(complete_table))]

complete_table2 <- complete_table[,-1]
completeCondition <- data.frame(condition=factor(c(rep("control", length(control_files)), rep("experimental", length(exp_files)))))
colnames(complete_table2) <- completeCondition$condition
dds <- DESeqDataSetFromMatrix(complete_table2, completeCondition, ~ condition)
dds <- DESeq(dds)

baseMeanPerLvl <- sapply( levels(dds$condition), function(lvl) rowMeans( counts(dds,normalized=TRUE)[,dds$condition == lvl] ) )

res <- results(dds, contrast = c("condition", "experimental", "control"))
func_results <- data.frame(res)
func_results$Function <- complete_table$Simplified_Function
func_results <- merge(func_results, baseMeanPerLvl, by="row.names")
func_results <- func_results[,c(1,2,8,9,3,4,5,6,7)]
colnames(func_results)[c(3,4)] <- c("controlMean", "experimentalMean")

plotMA(dds, ylim=c(-10,10), main="DESeq2 comparison of functional annotations")

sorted_func_results <- func_results[order(-func_results$baseMean),]
colnames(sorted_func_results)[1] <- "Function Name"
write.table(func_results, file = "DESeq_functional_output_with_md5s.tab", append = FALSE, quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)
```

The DESeq results file should now be saved with the file name of the last command just above, in a tab-delimited format.